---
title: "lncRNA"
author: "Bruno"
date: "2023-10-12"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries

```{r message=FALSE, warning=FALSE, include=FALSE}

library(dplyr)
library(tidyr)
library(forcats)
library(stringr)
library(ggplot2)
library(skimr)
library(here)

directorio <- here()
source("scriptlncrna.R")
```

You can check some preliminary analysis and summary statistics [here](./SomeStatistics.md)

## DIF Ratio

Histograms showing the diff-ratios of query (mRNA) and target (lncRNA) in both amputated/irradiated conditions.

The high proportion of minus-sense reads suggests accurate transcript assembly

**Opciones de conclusiones**

1.-The predominance of minus-sense reads strengthens the belief that the assembled transcripts are faithful representations of the original molecules.

2.-The substantial proportion of minus-sense reads strongly supports the conclusion that the assembled transcripts accurately represent the parental molecules.

3.-Finding mostly minus-sense reads increases our confidence that the assembled transcripts are indeed correctly assembled

4.-(La mayoría de lecturas están en el sentido negativo, lo cual da fiabilidad a que los transcritos que se ensamblaron en el sentido correcto.)

```{r}
#Histogram of the ratio of reads on the positive and the negative strand (ideal is -1)
ggplot(data = ampirr, aes(x = diff_ratio_ampirr.qer)) + geom_histogram() + 
  labs(x = "Difference ratio of reads", title = "Query mRNAs show the expected predominance of minus strand reads", subtitle = "Difference ratio taking into account both experiments", y = "Number of transcripts") + theme(plot.title = element_text(hjust = 0.5)) + xlim(-1,1)
ggplot(data = ampirr, aes(x = diff_ratio_ampirr.target)) + geom_histogram() + 
  labs(x = "Difference ratio of reads", title = "Antisense mRNAs show an even distribution of read strand", subtitle = "Difference ratio taking into account both experiments", y = "Number of transcripts") + theme(plot.title = element_text(hjust = 0.5)) + xlim(-1,1)

```

Expresión diferencial para las condiciones de amputación y radiación

# Padj significativa en los 4 cuadrantes con todos los datos

SelfBlast is a data frame that shows the alignment of a query transcript (mRNA) with a subject transcript (lncRNA). BLAST identifies some queries (mRNA) that have alignments with more than one subject (lncRNA). SelfBlast has 9,930 unique query (mRNA) transcript IDs.

The following data frames are based on SelfBlast plus the values of log2FoldChange and Padj. From the 80,000 entries, we kept only the with the entries padj \< 0.05.

## Irradiated

### mRNA (Query)

Filtering Padj \< 0.05 for mRNA/query

Number of rows: 223

| Variable           |  NA | complete_rate | unique |  mean |     p0 |   p25 |   p50 |
|:-------------------|----:|--------------:|-------:|------:|-------:|------:|------:|
| qseqid             |   0 |          1.00 |     25 |    NA |     NA |    NA |    NA |
| sseqid             |   0 |          1.00 |    217 |    NA |     NA |    NA |    NA |
| log2FoldChange.qer |   0 |          1.00 |     NA | -1.21 | -23.26 | -5.54 | -1.10 |
| padj.qer           |   0 |          1.00 |     NA |  0.01 |   0.00 | 0.004 |  0.01 |
| log2FoldChange.sub |   8 |          0.96 |     NA |  0.05 |  -3.48 | -0.46 |  0.11 |
| padj.sub           | 113 |          0.49 |     NA |  0.99 |   0.55 |  0.99 |  0.99 |

We have a total of 223 entries. Of these entries, 113 rows contain NAs in the subject Padj column (only the 49% of the entries have a value) and 8 NAs in the log2FoldChange subject column.

```{r}
#Script para el skim
#filter(mrna_deg, padj.qer.irr < 0.05) %>% skim_without_charts()
filter(mrna_deg, padj.qer.irr < 0.05) %>%
  ggplot(aes(x = log2FoldChange.sub.irr, y = log2FoldChange.qer.irr)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
labs(title = "mRNA diferential expressed when lncRNA not under Irradiation condition",
     x = "log2Fold change lncRNA",
     y = "log2Fold change mRNA")#, color = "red", linetype = "dashed" <- linea punteada)
```

Reminder that BLAST identified that some mRNA (query) transcripts align with more than one lncRNA (subject), hence the reason we observe many dots at the same level on the Y-axis. In this graphic, we observed dots where the mRNA (query) had significant DE (differential expression), but the pair, lncRNA/subject, did not. We might be interested in the mRNA TRINITY_DN5414_c2_g1_i3 that has a log2FoldChange of -23.261. We observed that TRINITY_DN5414_c2_g1_i3 aligns with 3 different lncRNAs.

PS: Notice how the axis X goes from -3 to 3, which means that the lncRNAs barely change and it wasn't significant.

### lncRNA (Subject)

Filtering Padj \< 0.05 for subject/lncRNA

── Data Summary ──────────────────────── Values\
Name Piped data Number of rows 101\
Number of columns 10

| Variable           |  NA | complete_rate | n_unique | mean |    p0 |   p25 |  p50 |
|:-------------------|----:|--------------:|---------:|-----:|------:|------:|-----:|
| qseqid             |   0 |          1.00 |       88 |   NA |    NA |    NA |   NA |
| sseqid             |   0 |          1.00 |       28 |   NA |    NA |    NA |   NA |
| log2FoldChange.qer |   2 |          0.98 |       NA | 0.37 | -2.17 | -0.19 | 0.13 |
| padj.qer           |  32 |          0.68 |       NA | 0.97 |  0.38 |  0.99 | 0.99 |
| log2FoldChange.sub |   0 |          1.00 |       NA | 0.61 | -6.93 | -1.15 | 1.03 |
| padj.sub           |   0 |          1.00 |       NA | 0.02 |  0.00 |  0.00 | 0.02 |

Filtering with Padj \< 0.05 for subject/lncRNA, we have a total of 101 entries. Of these entries, 32 rows contain NAs in the query Padj column (the 68% of entries have a value) and 2 NAs in the log2FoldChange subject column.

```{r}
# script skim
#filter(mrna_deg, padj.sub.irr < 0.05) %>% skim_without_charts()
filter(mrna_deg, padj.sub.irr < 0.05) %>%
  ggplot(aes(x = log2FoldChange.sub.irr, y = log2FoldChange.qer.irr)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
labs(title = "lncRNA diferential expressed",
     x = "log2Fold change lncRNA",
     y = "log2Fold change mRNA")

```

On the other hand, here we observed cases when the lncRNA values show significant differential expression, the log2FoldChange values for mRNA do not. The log2FoldChange values for lncRNA range between -7 and 7. Nonetheless, there could be some regulatory processes at play, as the lncRNA is overexpressed or underexpressed while the mRNA remains unchanged. Across these graphics, we may be particularly interested in cases where one transcript exhibits differential expression while the other does not.

## Amputated

### mRNA (query)

Filtering Padj \< 0.05 for query/mRNA

── Data Summary ──────────────────────── Values\
Name Piped data Number of rows **293**\
Number of columns 10

| variable           |  NA | complete_rate | n_unique |  mean |       p0 |   p25 |  p50 |
|:--------|--------:|--------:|--------:|--------:|--------:|--------:|--------:|
| qseqid             |   0 |          1.00 |       51 |    NA |       NA |    NA |   NA |
| sseqid             |   0 |          1.00 |      268 |    NA |       NA |    NA |   NA |
| log2FoldChange.qer |   0 |          1.00 |       NA | -0.89 |    -7.83 | -5.46 | 0.44 |
| padj.qer           |   0 |          1.00 |       NA |  0.01 |     0.00 | 0.003 | 0.01 |
| log2FoldChange.sub |  11 |          0.96 |       NA |  0.22 |    -3.02 | -0.32 | 0.01 |
| padj.sub           |  76 |          0.74 |       NA |  0.83 | 0.000024 |  0.75 | 0.93 |

A total of 293 entries, 11 rows with NA for log2FoldChange subject column, and 76 NAs for padj.

```{r}
#filter(mrna_deg, padj.qer.amp < 0.05) %>% skim_without_charts()

filter(mrna_deg, padj.qer.amp < 0.05) %>%
  ggplot(aes(x = log2FoldChange.sub.amp, y = log2FoldChange.qer.amp)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
labs(title = "mRNA diferential expressed",
     x = "log2Fold change lncRNA",
     y = "log2Fold change mRNA")
```

This graphic is interesting because we observe a log2FoldChange value of 22.9 for one lncRNA transcript, but in the data frame there is no corresponding Padj value. However, in this graphic, we are interested in the dots that have non-zero values on the Y-axis and values close to zero on the X-axis.

## lncRNA (subject)

Filtering Padj \< 0.05 for subject/lncRNA

── Data Summary ──────────────────────── Values\
Name Piped data Number of rows **149**\
Number of columns 10

| variable           |  NA | complete_rate | n_unique |  mean |    p0 |     p25 |   p50 |
|:--------|--------:|--------:|--------:|--------:|--------:|--------:|--------:|
| qseqid             |   0 |          1.00 |      134 |    NA |    NA |      NA |    NA |
| sseqid             |   0 |          1.00 |       59 |    NA |    NA |      NA |    NA |
| log2FoldChange.qer |   2 |          0.98 |       NA |  0.15 | -5.46 |   -0.35 | -0.05 |
| padj.qer           |  14 |          0.90 |       NA |  0.80 |  0.00 |    0.66 |  0.89 |
| log2FoldChange.sub |   0 |          1.00 |       NA |  0.48 | -7.77 |   -0.98 |  1.21 |
| padj.sub           |   0 |          1.00 |       NA | 0.016 |  0.00 | 0.00033 | 0.006 |

For this case, we have 149 entries and only 2 NAs for the log2FoldChange query column and 14 for the padj query column.

```{r}
#filter(mrna_deg, padj.sub.amp < 0.05) %>% skim_without_charts()
filter(mrna_deg, padj.sub.amp < 0.05) %>%
  ggplot(aes(x = log2FoldChange.sub.amp, y = log2FoldChange.qer.amp)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
labs(title = "lncRNA diferential expressed",
     x = "log2Fold change lncRNA",
     y = "log2Fold change mRNA")
```

Here we observe significant values for lncRNAs (subject). This graphic includes lncRNAs for which BLAST identified alignments with more than one mRNA.

**ANOTACIONES PARA DEG**

## Casos de Alinieamiento

Descripción de casos:

-caso 1 El alineamiento es solamente dentro del CDS

-caso 2 El alineamiento es tan largo que cae en las 3 zonas (5UTR, CDS y 3UTR)

-caso 3 El alineamiento empieza en la zona 5UTR y termina dentro de CDS

-caso 4 El alineamiento empieza en la zona de CDS y termina en 3UTR

-caso 5 El alineamiento es solamente en la zona 5UTR

-caso 6 El alineamiento es solamente en la zona 3UTR

```{r warning=FALSE}
prot %>% ggplot(aes(x = caso)) + geom_histogram(stat = "count") + labs(title = "Most of the alignments are only in the 3'UTR area" ,subtitle = "Cases distribution")
```

# Código para histogramas

Histogramas Cuando el alineamiento es en la zona de CDS, la mayoría de alineamientos en la zona de CDS se alinea con el 50% del

```{r warning=FALSE}
prot %>% filter(CDSporcentaje>0) %>%   ggplot(aes(x = CDSporcentaje)) + geom_histogram()
```

Cuando el alineamiento es en la zona 3UTR, la gran mayoría de veces se alinea con todo el largo del área 3UTR

```{r warning=FALSE}
prot %>% filter(UTR3porcentaje>0) %>%   ggplot(aes(x = UTR3porcentaje)) + geom_histogram()
```

Cuando el alineamiento es en la zona 5UTR, la mayoría de veces el alineamiento es en todo el largo del área 5UTR

```{r warning=FALSE}
prot %>% filter(UTR5porcentaje>0) %>%   ggplot(aes(x = UTR5porcentaje)) + geom_histogram()
```

Correlación positiva para el caso 1 (el alineamiento es únicamente en la zona CDS)

**ANOTACIONES PARA LOS CASOS**

Es importante destacar que de los datos originales, cerca de 80mil, se hizo un "unique" para quitar los valores repetidos, el data frame paso a ser de 40mil datos, luego se filtró para quitar las filas que tenía NA en la columna de "inicio de proteina" y "final de proteína" es decir que no tenian las coordenadas de la proteína. Con eso nos quedamos con 25mil datos. Se hizo un último filtro para quedarnos solo con las proteínas que están en sentido. Es decir que el inicio de la proteína sea menor al final de la proteína.

## Explicación sobre los DF

Descripción del trabajo

**DF de expresión diferencial**:

-results_full_amp_degs

-results_full_irr_degs

**DF resultados del blast con transcritos antisentido**:

-swissProt_self_blastAStranscripts

**DF con anotaciones de BLASTX, prot ID, prot corrds, BLASTP, ontología, nombre de los genes, organismos en el que se encuentra, length, etc** :

-dlaevis_assembly_uniprt

## Data frames creadas:

\-**ampirr**: 32 columnas, 79mil filas Tomando como base el DF swissProt, se hizzo un left join de los DF de DEG. Además se hizo otro left join con las longitudes

ampirr tiene 79mil entradas porque a veces un transcrito de query, hace match con diferentes subjects. En total hay 9930 queries diferentes.

**-longitudes:** DF con las longitudes totales de los transcritos, esto se hizo utilizando un archivo fasta y utilizando grep

\-**X** El data frame x se hizo tomando como base ampirr y juntando los datos de log2FoldChange y padj de ambas condiciones.

\-**mrna_deg** El data frame mrna_deg se hizo tomando como base ampirr y juntando los datos de log2FoldChange y padj de ambas condiciones. Después

\-**Filtrados**:32 collumnas, 23mil filas. Tomando como base el DF ampirr, se hizo un filtro para seleccionar a los transcritos mayores a 300. El df filtrado contiene 23639 entradas, es decir, se quedarón fuera 50mil datos aprox

\-**qer_sub_diff**:10 coulmnas, 23mil filas. Tomando como base filtrados. Se hicieron left joins para agregar los datos de log2FoldChange y padj para qseqid y sseqid de ambas condiciones (amp y irr)

\-**medianos**: esta tabla tiene los 50mil datos que fueron excluidos en el DF filtrados, es decir, aquí hay transcritos de entres 200 y 300 nucleotidos.

\-**diff_rati_olvidados**: al DF olvidados, se le agregó los datos de log2FoldChange Gráficas con DEG de los olvidados con valores de padj menores a 0.05

**prot**: Se seleccionó unicamente las columnas de trasncript_id y prot coords del DF dlaevis_assembly_uniprt. Con código se separo prot coords en dos columnas (prot_start y prot_end). Se eliminaron las filas con NA en prot coord Se hizo un left join en prot con filtrados para agregar los datos de "qseqid, sseqid, length, qstart, qend,sstart, send" Se filtró para tener únicamente los datos en el que van en sentido Se agregaron columnas con los datos de CDS_length y URT_3 Se hicieron las columnas de los casos, porcentaje de zona cubierto, y porcentaje del lnc en cada zona.
